One-loop PDF of the average Ising magnetization in 4 dimensions.

### 2020-03-22 paper/paper.tex
Paper checkpoint

\begin{figure}
\begin{center}
\includegraphics[scale=0.75]{binder_cumulant.png}
\end{center}
\caption{\label{fig:binder_cumulant} Blah Blah}
\end{figure}

@@

In Monte Carlo simulations, the qualitative behavior of the distribution of the
magnetization is often characterized by measuring the so-called Binder
cumulant:
\begin{equation}
    U = 1 - \frac{\left\langle \bar \phi^4 \right \rangle}
    {3\left\langle \bar \phi^2 \right \rangle^2},
\end{equation}
which is constructed to be independent of the overall scale of $\bar{\phi}$,
and to be zero if $\bar{\phi}$ is normally distributed. From
(\ref{eq:log_pdf_critical_point}) we conclude that, on a 4-Torus, at the
critical point\footnote{ Incidentally, we note that this result is not in
contraddiction with the claim in the original paper that the critical Binder
cumulant in 4 dimensions is zero, because in the original paper the cumulant is
not computed from the total magnetization, but rather from the average
magnetization of blocks of spins much smaller than the overall system size.}:
\begin{equation}
    \label{eq:critical_binder_cumulant}
    U = 1 - \frac{4\,\Gamma\left(\frac{5}{4}\right)^2}
    {3\,\Gamma\left(\frac{3}{4}\right)^2} 
    = 0.27052\ldots
\end{equation}

In fig.~\ref{fig:binder_cumulant}, we show a comparison of the Binder cumulant
computed from (\ref{eq:log_pdf}) near the critical point, and the results of
Monte Carlo simulation of the Ising model.

@@

We now describe briefly how the result (\ref{eq:log_pdf}) is obtained. First of
all we observe that the average magnetization $\bar{\phi}$ is proportional to
the zero momentum mode of the field $\phi$:
\begin{equation}
    \phi(x) = \frac{1}{V} \sum_{n \in \mathbb{Z}^4} \phi_{n} 
    e^{i \frac{n\cdot x}{R}}\,;\quad \bar{\phi} = \frac{\phi_{0}}{V}\,.
\end{equation}

@@

The summation $I_1$ can be simplified substantially using the following trick:
\begin{equation}
\begin{split}
    I_1 &= \frac{1}{V} \int_{s}^{\infty} \dd t \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    e^{-t(r + n^2 R^{-2})}\\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\left(\sum_{m = -\infty}^{\infty} e^{-t m^2 R^{-2}}\right)^4 - 1\right) \\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\theta\!\left(t R^{-2}\right)^4 - 1\right)\,,
\end{split}
\end{equation}
where:\footnote{The function $\theta$ is in fact a Jacobi theta function, but we
won't need any of its special properties beyond its asymptotic behavior, which
can be easily obtained from the definition.}
\begin{equation}
    \theta(z) \equiv \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
\end{equation}

We now extract the divergent part of $I_1$ by subtracting under the integral a
function that has the same asymptotic behavior as the integrand for $t \to 0$,
but whose integral can be computed in closed form. In order to do that, we need
the asymptotic behavior of $\theta(z)$ for $z\to0$, which can be obtained from
its definition using the Euler-Mclaurin formula:
\begin{equation}
    \theta(z) \sim \sqrt{\frac{\pi}{z}} + o(z^k) \quad\mathrm{for}\, z\to0\,.
\end{equation}

Thus we have:
\begin{equation}
\begin{split}
    I_1 &= \frac{1}{16\pi^2}\left(\int_{s}^{\infty}\dd t\, e^{-t \mu^2} t^{-2}\left(1 + t (\mu^2 - r)\right)
    +\frac{1}{R^2} f(r R^2, \mu R)\right)\\
    &= \frac{1}{16\pi^2}\left(\frac{1}{s} + r\left(\log s\mu^2 + \gamma_{E}\right) - 
    \mu^2 +\frac{1}{R^2} f(r R^2, \mu R)\right)\,,
\end{split}
\end{equation}
where $\mu$ is a renormalization scale that can be chosen at will, and:
\begin{equation}
    f(rR^2,\mu R) = \int_{0}^{\infty} \dd z\ e^{-z r R^2} \left(\frac{\theta(z)^4 - 1}{\pi^2} - 
    \frac{1 + z\left(\mu^2 - r\right)R^2}{t^2}e^{-z\left(\mu^2 - r\right)R^2}\right).
\end{equation}

\begin{equation}
\label{eq:f_functions}
f_1 = 
\end{equation}

>>>>

\begin{figure}
\begin{center}
\includegraphics[scale=0.75]{binder_cumulant.png}
\end{center}
    \caption{\label{fig:binder_cumulant} Binder cumulant of the total
    magnetization of the 4D Ising model. The left plot gives a broad picture,
    the right plot shows the critical region close-up.  The shaded regions
    display the one-sigma confidence intervals obtained from Monte Carlo
    simulation. The center lines are obtained from a 3-parameter fit of the
    perturbative result (\ref{eq:log_pdf}). The horizontal dashed line
    indicates the critical value (\ref{eq:critical_binder_cumulant}). The fit
    has $J_c = 0.1496938$, $J - J_c = -0.027\cdot r$, $g = 0.43$, with $r$, $g$
    defined at the renormalization scale $L = 2\pi R = 8$. }
\end{figure}

@@

In Monte Carlo simulations, the qualitative behavior of the distribution of the
magnetization is often characterized by measuring the so-called Binder
cumulant \cite{Binder1981}:
\begin{equation}
    U = 1 - \frac{\left\langle \bar \phi^4 \right \rangle}
    {3\left\langle \bar \phi^2 \right \rangle^2},
\end{equation}
which is constructed to be independent of the overall scale of $\bar{\phi}$,
and to be zero if $\bar{\phi}$ is normally distributed. From
(\ref{eq:log_pdf_critical_point}) we conclude that, on a 4-Torus, at the
critical point\footnote{ Incidentally, we note that this result is not in
contraddiction with the claim in the original paper \cite{Binder1981} that the
critical Binder cumulant in 4 dimensions is zero, because in the original paper
the cumulant is not computed from the total magnetization, but rather from the
average magnetization of blocks of spins much smaller than the overall system
size.}:
\begin{equation}
    \label{eq:critical_binder_cumulant}
    U = 1 - \frac{4\,\Gamma\left(\frac{5}{4}\right)^2}
    {3\,\Gamma\left(\frac{3}{4}\right)^2} 
    = 0.27052\ldots
\end{equation}

In fig.~\ref{fig:binder_cumulant}, we show a comparison of the Binder cumulant
computed from (\ref{eq:log_pdf}) near the critical point, and the results of
Monte Carlo simulation of the $4D$ Ising model. The agreement is excellent except
for the smallest system size $L = 8$, probably because of contributions from
irrelevant operators. Notice how slowly the finite size Binder cumulant
approaches the asymptotic value (\ref{eq:critical_binder_cumulant}).

@@

We now describe briefly how the result (\ref{eq:log_pdf}) is obtained. The
perturbative approach is very similar to the computation of the effective
action, as in \textit{e.g.} \cite{WeinbergEffectiveAction}, except that we are
interested in the whole probability distribution of $\bar{\phi}$, instead of
just the expected value. The main difficulty lies in evaluating the loop
integrals at finite size.

First of all we observe that the average magnetization
$\bar{\phi}$ is proportional to the zero momentum mode of the field $\phi$:
\begin{equation}
    \phi(x) = \frac{1}{V} \sum_{n \in \mathbb{Z}^4} \phi_{n} 
    e^{i \frac{n\cdot x}{R}}\,;\quad \bar{\phi} = \frac{\phi_{0}}{V}\,.
\end{equation}

@@

The summation $I_1$ can be simplified substantially using the following trick:
\begin{equation}
\begin{split}
    I_1 &= \frac{1}{V} \int_{s}^{\infty} \dd t \sum_{n\in\mathbb{Z}^4\setminus\{0\}} 
    e^{-t(r + n^2 R^{-2})}\\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\left(\sum_{m = -\infty}^{\infty} e^{-t m^2 R^{-2}}\right)^4 - 1\right) \\
    & = \frac{1}{V} \int_{s}^{\infty} \dd t\ e^{-t r}
        \left(\theta\!\left(t R^{-2}\right)^4 - 1\right)\,,
\end{split}
\end{equation}
where:\footnote{The function $\theta$ is in fact a Jacobi theta function, but we
won't need any of its special properties beyond its asymptotic behavior, which
can be easily obtained from the definition.}
\begin{equation}
    \label{eq:theta}
    \theta(z) \equiv \sum_{m = -\infty}^{\infty} e^{-m^2 z}\,.
\end{equation}

We now extract the divergent and finite parts of $I_1$ as $s\to 0$ by
subtracting under the integral a function that has the same asymptotic behavior
as the integrand for $t \to 0$, but whose integral can be computed in closed
form. In order to do that, we need the asymptotic behavior of $\theta(z)$ for
$z\to0$, which can be obtained from its definition using the Euler-Mclaurin
formula:
\begin{equation}
    \theta(z) \sim \sqrt{\frac{\pi}{z}} + o(z^k) \quad\mathrm{for}\, z\to0\,.
\end{equation}

Thus we have:
\begin{equation}
\begin{split}
    I_1 &= \frac{1}{16\pi^2}\left(\int_{s}^{\infty}\dd t\, e^{-t \mu^2} t^{-2}\left(1 + t (\mu^2 - r)\right)
    +\frac{1}{R^2} f(r R^2, \mu R) + O(s)\right)\\
    &= \frac{1}{16\pi^2}\left(\frac{1}{s} + r\left(\log s\mu^2 + \gamma_{E}\right) - 
    \mu^2 +\frac{1}{R^2} f(r R^2, \mu R) + O(s)\right)\,,
\end{split}
\end{equation}
where $\mu$ is a renormalization scale that can be chosen at will, and:
\begin{equation}
    \label{eq:f_definition}
    f(\bar{r},\bar{\mu}) = \int_{0}^{\infty} \dd z\ e^{-z \bar{r}} \left(\frac{\theta(z)^4 - 1}{\pi^2} - 
    \frac{1 + z\left(\bar{\mu}^2 - \bar{r}\right)}{z^2}e^{-z\left(\bar{\mu}^2 - \bar{r}\right)}\right).
\end{equation}

We set the counterterms to:
\begin{align}
    &\delta r = -\frac{1}{32\pi^2}\left(\frac{1}{s} + r\left(\log s \mu^2 + \gamma_E - 1\right)\right)\,,\\
    &\delta u = -\frac{3}{32\pi^2}\left(\log s\mu^2 +\gamma_E + 1\right)\,,
\end{align}
from which the renormalization group equations (\ref{eq:renormalization_group})
can be obtained, and we have:
\begin{equation}
\begin{split}
    -\log p(\bar{\phi}) =& \mathrm{const}+ V \Bigg( \\
    &\frac{1}{2} \bar{\phi}^2\left( r + \frac{u}{32\pi^2}  \left(\left(\frac{1}{R^2} f(rR^2, \mu R) + r - \mu^2\right)\right) + O(u^2)\right) + \\
    &\frac{u}{24} \bar{\phi}^4 \left(1 + \frac{3u}{32\pi^2} f^{(1, 0)}(rR^2, \mu R) + O(u^2)\right) + \\
    &\frac{u^3}{48}\bar{\phi}^6\left(\frac{R^2}{32\pi^2}f^{(2, 0)}(rR^2, \mu R) + O(u)\right) + O(u^4 \bar{\phi}^8) \Bigg)\,,
\end{split}
\end{equation}
from which (\ref{eq:log_pdf}) is obtained by setting $g = \frac{3 u}{16
\pi^2}$ and $\mu^2 = r + R^{-2}$. This last choice is motivated as follows. The
subtraction in (\ref{eq:f_definition}) is similar to:
\begin{equation}
    \int_0^{\infty} \dd z\, \frac{e^{-a z} - e^{-b z}}{z} = \log \frac{b}{a}\,;
\end{equation}
when the asymptotic behavior of the two terms for large $z$ is not well
matched, the integral becomes large in magnitude, making the perturbative
expansion less reliable. With the choice $\mu^2 = r + R^{-2}$, both terms in
(\ref{eq:f_definition}) have the same asymptotic behavior $\sim e^{-z(\bar{r} +
1)}$, thus avoiding the large log problem over the widest possible range of
parameters.

For completeness, let us display explicitly the functions $f_i$ that
parametrize (\ref{eq:log_pdf}):
\begin{align}
\label{eq:f_functions}
    &f_1(\bar r) = f(\bar r, \sqrt{\bar r + 1}) = 
    \int_{0}^{\infty} \dd z\ e^{-z \bar{r}} \left(\frac{\theta(z)^4 - 1}{\pi^2} - 
    \frac{(1 + z) e^{-z}}{z^2}\right)\,,\\
    &f_2(\bar r) = f^{(1, 0)}(\bar r, \sqrt{\bar r + 1}) = 
    \int_{0}^{\infty} \dd z\ e^{-z \bar{r}} z \left(\frac{\theta(z)^4 - 1}{\pi^2} - 
    \frac{e^{-z}}{z^2}\right)\,,\\
    &f_3(\bar r) = f^{(2, 0)}(\bar r, \sqrt{\bar r + 1}) = 
    \int_{0}^{\infty} \dd z\ e^{-z \bar{r}} z^2 \frac{\theta(z)^4 - 1}{\pi^2}\,,
\end{align}
and $\theta$ is given by (\ref{eq:theta}).

\begin{thebibliography}{9}
\bibitem{Binder1981}
K. Binder,
\textit{Critical Properties from Monte Carlo Coarse Graining and Renormalization},
\href{https://doi.org/10.1103/PhysRevLett.47.693}{Phys. Rev. Lett. \textbf{47} (1981) 693}.

\bibitem{WeinbergEffectiveAction}
S. Weinberg,
\textit{The quantum theory of fields, Vol. 2}, Cambridge University Press, (1996) pg. 63.
\end{thebibliography}
